Supplemental Information: Continuous developmental changes in word recognition support language learning across early childhood

### 2025-03-14 paper/paper.Rmd
SI

Our first question was about the functional form of the relationships between age, speed, and accuracy (see SI: Correlations for raw correlations between variables). We began by fitting linear mixed effects models predicting speed and accuracy on each trial across the full dataset with random slopes of child age by study and random intercepts by participant. We compared models that included both long and short accuracy windows, as well as logarithmic and linear effects of age, and logarithmic and linear transformations of RT. The best fitting model of accuracy predicted long window accuracy as a function of the logarithm of age,^[This longer window yielded overall higher cross-trial reliability as well, and so we report long window results in the main text. All findings reported here hold for both window sizes, however (see SI: Short Window Analysis).] and the best fitting model of speed predicted log RT as a function of log age as well (see SI: Functional Form Model Comparison).

@@

>>>>

Our first question was about the functional form of the relationships between age, speed, and accuracy (see SI: Correlations for raw correlations between variables). We began by fitting linear mixed effects models predicting speed and accuracy on each trial across the full dataset with random slopes of child age by study and random intercepts by participant. We compared models that included both long and short accuracy windows, as well as logarithmic and linear effects of age, and logarithmic and linear transformations of RT. The best fitting model of accuracy predicted long window accuracy as a function of the logarithm of age; the best fitting model of speed predicted log RT as a function of log age as well (see SI: Functional Form Model Comparison). Because long window accuracies were more correlated with other variables and showed clearer age gradients, we focus on these in our analysis.

@@

We excluded trials entirely if they were missing data on more than 50% of timepoints, and excluded RTs if they were based on fewer than 50% of timepoints in the short analytic window (200 -- 2000 ms). We also removed RTs shorter than 367 ms, as these were unlikely to be generated based on the specific linguistic stimulus. We then excluded participants from the analysis if they contributed fewer than four accuracy measurements or fewer than 2 reaction time measurements. At the participant level, these steps together led to `r round(mean(is.na(d_sub$rt)), 2)`% missingness for RTs and `r round(mean(is.na(d_sub$long_window_accuracy)), 2)`% missingness for long window accuracies.

### 2025-03-14 paper/supplemental_information.Rmd
SI

---
title: "Supplemental Information"
output: pdf_document
header-includes:
  - \usepackage{threeparttablex}
  - \usepackage{booktabs}
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(here)
source(here("helper","common.R"))
```

```{r, echo=FALSE}
knitr::purl(here("paper","paper.Rmd"), documentation = 0)
suppressMessages(source(here("paper.R")))
```

@@

```{r, fig.env="figure*"}
ggplot(d_sub, aes(x = age))+ 
  geom_histogram(binwidth = 1) + 
  facet_wrap(~dataset_name, scale = "free_y")
```

```{r, fig.cap="Number of administrations for datasets with repeated measurements."}

@@

ggplot(d_sub_long, 
       aes(x = delta_t, fill = dataset_name)) +
  geom_histogram(binwidth = 1) + 
  # facet_wrap(~dataset_name) + 
  xlab("Age since first test")
```
# Test-Retest Reliability

@@

d_reliability <- d_long |>
  filter(time_since_t0 <= MONTH_CUTOFF, 
         admin_num <= 2) |>
  pivot_wider(id_cols = c("subject_id","dataset_name"), 
              names_from = "admin_num",
              values_from = c("log_rt", "log_rt_var", "long_window_accuracy", 
                              "long_window_acc_var", "prod", "comp"))

dataset_reliabilities <- d_reliability |>
  group_by(dataset_name) |>
  summarise(rt = cor(log_rt_1, log_rt_2, use = "pairwise.complete.obs"),
            acc = cor(long_window_accuracy_1, 
                          long_window_accuracy_2, use = "pairwise.complete.obs"))

# note this is not correct for CDIs because some CDIs can get grouped with two admins, inflating correlations
global_reliabilities <- d_reliability |>
  ungroup() |>
  summarise(rt = cor(log_rt_1, log_rt_2, use = "pairwise.complete.obs"),
            acc = cor(long_window_accuracy_1, 
                          long_window_accuracy_2, use = "pairwise.complete.obs"))
```

We examined test-retest reliability for our primary variables of interest by calculating Pearson correlations between pairs of administrations given no more than three months apart. Test-retest correlations were significant but relatively modest ($\rho_{acc} = `r round(global_reliabilities[2], 3)`$, $\rho_{rt} =  `r round(global_reliabilities[1],3)`$). These reliabilities were biased downwards by three factors, however. First, longitudinal assessments sometimes use variable items between testing sessions, leading to item-related variance in measurement. Second, even three months can lead to substantial change in some children's language abilities, thus correlations are attenuated by true change as well as measurement error. Third, longitudinal data in the dataset come primarily from the youngest children and hence are likely to show overall higher measurement error due to variability in children's behavior and an overall lower number of trials.

@@

```{r, results = "asis"}

@@

```{r results="asis"}
papaja::apa_table(acc_mods_lmer_summary)
```

```{r results="asis"}
papaja::apa_table(rt_mods_lmer_summary)
```

With our current dataset we cannot specifically determine whether within-individual patterns of change conform to power law or exponential forms [@kail1991,@heathcote2000], but regardless we see clear patterns of developmental change related to skill acquisition.

@@

# Short Window Analysis

Need to redo all figures using short windows ugh.

>>>>

---
title: "Supplemental Information: Continuous developmental changes in word recognition support language learning across early childhood"
output: 
  pdf_document:
    fig_caption: yes
    keep_tex: yes
header-includes:
  - \usepackage{threeparttablex}
  - \usepackage{booktabs}
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE)
library(here)
source(here("helper","common.R"))
```

```{r, echo=FALSE, cache=TRUE, message=FALSE, warning=FALSE}
paper <- suppressMessages(knitr::purl(here("paper","paper.Rmd"), documentation = 0))
suppressMessages(source(paper))
```

@@

Figure \ref{fig:dataset} gives the age distribution for each separate dataset.

```{r dataset, fig.cap="\\label{fig:dataset} Age distribution for each dataset, using three-month bins.", fig.pos="h!"}
ggplot(d_sub, aes(x = age))+ 
  geom_histogram(binwidth = 3) + 
  facet_wrap(~dataset_name, scale = "free_y") +
  xlab("Number of participants") + 
  ylab("Age (months)")
```

Figure \ref{fig:longitudinal} shows the distribution of measurement intervals for longitudinal studies within the dataset.

```{r longitudinal, fig.cap="\\label{fig:longitudinal} Number of administrations for datasets with repeated measurements, plotted by dataset.", fig.pos="h!"}

@@

ggplot(d_sub_long, 
       aes(x = delta_t, fill = dataset_name)) +
  geom_histogram(binwidth = 1) + 
  scale_fill_solarized(name = "Dataset") +
  theme(legend.position = "bottom") + 
  # facet_wrap(~dataset_name) + 
  xlab("Age since first test") + 
  ylab("Number of participants")
```

# Test-Retest Reliability

@@

d_reliability <- d_long |>
  filter(time_since_t0 <= MONTH_CUTOFF, 
         admin_num <= 2) |>
  pivot_wider(id_cols = c("subject_id","dataset_name"), 
              names_from = "admin_num",
              values_from = c("log_rt", "long_window_accuracy", 
                              "short_window_accuracy"))

dataset_reliabilities <- d_reliability |>
  group_by(dataset_name) |>
  summarise(rt = cor(log_rt_1, log_rt_2, use = "pairwise.complete.obs"),
            acc_long = cor(long_window_accuracy_1, 
                           long_window_accuracy_2, use = "pairwise.complete.obs"),
            acc_short = cor(short_window_accuracy_1, 
                            short_window_accuracy_2, use = "pairwise.complete.obs"))

global_reliabilities <- d_reliability |>
  ungroup() |>
  summarise(rt = cor(log_rt_1, log_rt_2, use = "pairwise.complete.obs"),
            acc_long = cor(long_window_accuracy_1, 
                           long_window_accuracy_2, use = "pairwise.complete.obs"),
            acc_short = cor(short_window_accuracy_1, 
                            short_window_accuracy_2, use = "pairwise.complete.obs"))
```

We examined test-retest reliability for our primary variables of interest by calculating Pearson correlations between pairs of administrations given no more than three months apart. Test-retest correlations were significant but relatively modest: $\rho_{long window acc} = `r round(global_reliabilities[2], 3)`$, $\rho_{short window acc} = `r round(global_reliabilities[3], 3)`$, $\rho_{rt} =  `r round(global_reliabilities[1],3)`$. These reliabilities were biased downwards by three factors, however. First, longitudinal assessments sometimes use variable items between testing sessions, leading to item-related variance in measurement. Second, even three months can lead to substantial change in some children's language abilities, thus correlations are attenuated by true change as well as measurement error. Third, longitudinal data in the dataset come primarily from the youngest children and hence are likely to show overall higher measurement error due to variability in children's behavior and an overall lower number of trials.

@@

Table \ref{tab:corrs} shows pairwise correlations between the primary variables of interest in the dataset.

\begin{table}[tbp]
\begin{center}
\begin{threeparttable}
\begin{tabular}{llllllllll}
\toprule
& \multicolumn{1}{l}{age} & \multicolumn{1}{l}{log age} & \multicolumn{1}{l}{RT} & \multicolumn{1}{l}{log RT} & \multicolumn{1}{l}{long acc} & \multicolumn{1}{l}{short acc } & \multicolumn{1}{l}{prod} & \multicolumn{1}{l}{comp}\\
\midrule
age &  1.00 &  &  &  &  &  &  & \\
log age  & 0.98 & 1.00 &  &  &  &  &  &\\
rt  & -0.33 & -0.35 & 1.00 &  &  &  &  & \\
log rt  & -0.34 & -0.36 & 0.96 & 1.00 &  &  &  & \\
long window accuracy  & 0.44 & 0.48 & -0.48 & -0.46 & 1.00 &  &  & \\
short window accuracy & 0.38 & 0.43 & -0.62 & -0.61 & 0.82 & 1.00 &  & \\
production vocabulary & 0.72 & 0.70 & -0.31 & -0.33 & 0.51 & 0.45 & 1.00 & \\
comprehension vocabulary & 0.42 & 0.42 & -0.25 & -0.24 & 0.24 & 0.24 & 0.59 & 1.00\\
\bottomrule
\end{tabular}
\end{threeparttable}
\end{center}
\caption{Pairwise correlations between primary variables of interest. \label{tab:corrs}}
\end{table}

```{r corrs, results = "asis", include = FALSE}

@@

Table \ref{tab:accmods} shows model comparison measures between for different models of the functional form of the relationship between accuracy and age and Table \ref{tab:rtmods} shows the same for reaction time. Age gradients are estimated substantially better with long window accuracies. Note that there are a greater number of observations for short window accuracies due to less missing data. We speculate that, on average, more participants looked away from the screen towards the end of trials, leading to a greater number of exclusions of long window trials based on the 50% criterion. Note that the total percentage of trials excluded is still small for both measures: `r round(mean(is.na(d_trial$long_window_accuracy)), 2)`% for long window accuracy and `r round(mean(is.na(d_trial$short_window_accuracy)), 2)`% for short window accuracy.

\begin{table}[tbp]
\begin{center}
\begin{threeparttable}
\begin{tabular}{lllllllll}
\toprule
n. obs & \multicolumn{1}{c}{sigma} & \multicolumn{1}{c}{logLik} & \multicolumn{1}{c}{AIC} & \multicolumn{1}{c}{BIC} & \multicolumn{1}{c}{REMLcrit} & \multicolumn{1}{c}{df.residual} & \multicolumn{1}{c}{model} & \multicolumn{1}{c}{$r_2$}\\
\midrule
48354 & 0.28 & -8,338.83 & 16,691.65 & 16,753.16 & 16,677.65 & 48347 & Long window, linear age & 0.14\\
48354 & 0.28 & -8,305.39 & 16,624.78 & 16,686.28 & 16,610.78 & 48347 & Long window, log age & 0.12\\
50244 & 0.31 & -12,486.60 & 24,987.21 & 25,048.98 & 24,973.21 & 50237 & Short window, linear age & 0.09\\
50244 & 0.31 & -12,458.77 & 24,931.54 & 24,993.31 & 24,917.54 & 50237 & Short window, log age & 0.08\\
\bottomrule
\end{tabular}
\end{threeparttable}
\end{center}
\caption{Model comparison metrics for different functional forms of the relationship between accuracy and age. \label{tab:accmods}}
\end{table}

\begin{table}[tbp]
\begin{center}
\begin{threeparttable}
\begin{tabular}{lllllllll}
\toprule
n. obs & \multicolumn{1}{c}{sigma} & \multicolumn{1}{c}{logLik} & \multicolumn{1}{c}{AIC} & \multicolumn{1}{c}{BIC} & \multicolumn{1}{c}{REMLcrit} & \multicolumn{1}{c}{df.residual} & \multicolumn{1}{c}{model} & \multicolumn{1}{c}{$r_2$}\\
\midrule
18940 & 0.45 & -12,414.69 & 24,843.38 & 24,898.32 & 24,829.38 & 18933 & Log RT, linear age & 0.21\\
18940 & 0.45 & -12,393.27 & 24,800.53 & 24,855.48 & 24,786.53 & 18933 & Log RT, log age & 0.21\\
18940 & 570.49 & -147,644.43 & 295,302.87 & 295,357.81 & 295,288.87 & 18933 & Linear RT, linear age & 0.18\\
18940 & 570.02 & -147,622.67 & 295,259.34 & 295,314.29 & 295,245.34 & 18933 & Linear RT, log age & 0.18\\
\bottomrule
\end{tabular}
\end{threeparttable}
\end{center}
\caption{Model comparison metrics for different functional forms of the relationship between RT and age. \label{tab:rtmods}}
\end{table}

Note that with our current dataset we cannot specifically determine whether within-individual patterns of change conform to linear, power law, or exponential developmental patterns (Heathcote, Brown, \& Mewhort, 2000). Our current results apply to the form of the age gradient as opposed to the form of any individual's pattern of developmental change.

```{r accmods, results="asis", include = FALSE}
papaja::apa_table(acc_mods_lmer_summary)
```

```{r results="asis", include = FALSE}
papaja::apa_table(rt_mods_lmer_summary)
```

@@

<!-- # Short Window Analysis -->

<!-- Need to redo all figures using short windows ugh.  -->
# References

Heathcote, A., Brown, S., \& Mewhort, D. J. K. (2000). The power law repealed: The case for an exponential law of practice. Psychonomic Bulletin & Review, 7(2), 185–207.
